---
title: "Kwoma"
author: "Bill Mizgerd"
bibliography: kmo.bib
output: html_document
---

Last Updated: 2019-07-24

# Background

**Language Family**: Sepik / Middle Sepik / Nukuma / Kwoma

* Also referred to as Washkuk, Kwoma is spoken in 12 villages within the East Sepik province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar    Postalveolar     Palatal     Velar     Glottal
Stops                       "ᵐb"      "t  ⁿd"         ""             ""       "k  ᵑɡ"      "ʔ"
Affricates                  ""           ""        "tʃ  ⁿdʒ"         ""         ""         ""
Fricatives                 "ɸ  β"       "s"           "ʃ"            ""         ""         "h"
Nasals                      "m"         "n"           ""             "ɲ"        ""         ""
Flaps                       ""          "ɾ"           ""             ""         ""         ""
Approximants                "w"         ""            ""             "j"        ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Phonemes that contain a diacritic /ᵐ/, /ⁿ/, or /ᵑ/ are prenasalized.", general_title = "")
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"         "u"
Mid       "e"         ""          "o"
""        "ɛ"         ""          ""
Low       ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"            "/a/"
"b"            "/ᵐb/"    
"d"            "/ⁿd/"  
"e"            "/e/"  
"g"            "/ᵑɡ/"  
"h"            "/h/"  
"i"            "/i/"  
"j"            "/ⁿdʒ/"  
"k"            "/k/"  
"m"            "/m/"  
"n"            "/n/"   
"o"            "/o/"  
"p"            "/ɸ/"  
"r"            "/ɾ/"  
"s"            "/s/"    
"t"            "/t/"  
"u"            "/u/"  
"v"            "/β/"  
"w"            "/w/"  
"y"            "/j/"  
"\'"            "/ʔ/"  
**Digraph**     ""
"ch"           "/tʃ/"
"ny"           "/ɲ/"
"sh"           "/ʃ/"
"ay"           "/ai/"  
"ee"           "/ɛ/"  
"ey"           "/ei/"  
"ii"           "/ɨ/"  
"iy"           "/i/"  
"ow"           "/ou/"  
"uw"           "/u/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* The suffix /-ta/ spirantizes to /-tʃa/ after certain morphemes [@KmoKooyers1974, p. 9].
* The suffix /-tawa/ collapses to one of four shortened versions, depending on preceding morphemes [@KmoKooyers1974, p. 10].
* Monosyllabic /a/-final verb stems raise to [ɛ] before the suffix /-tʃi/ (ibid.).
* The imperative marker /na/ raises to [ne] before /i/ (ibid.).

# References
