---
title: "Maltese"
author: "Bill Mizgerd"
bibliography: mt.bib
output: html_document
---

Last Updated: 2018-10-21

# Background

**Language Family**: Afro-Asiatic / Semitic / Central Semitic / Arabic / Western Arabic / Maltese

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar    Postalveolar   Palatal     Velar     Pharyngeal     Glottal
Stops                      "p  b"     "t̪  d̪"      ""           ""           ""       "k  ɡ"         ""           "ʔ"
Affricates                   ""        ""      "ts  dz"     "tʃ  dʒ"        ""         ""           ""           ""
Fricatives                 "f  v"      ""       "s  z"       "ʃ  ʒ"         ""         ""           "ħ"          ""
Nasals                       "m"       ""         "n"          ""           ""         ""           ""           ""
Trills                       ""        ""         "r"          ""           ""         ""           ""           ""
Approximants                 ""        ""         "l"          ""           "j"        "w"          ""           ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Pharyngeal", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowel length is contrastive.
* Authors seem to vary in terms of the vowel inventory:
  - (1996) HUME:             /a/ /e/ /i/ /o/ /u/  
  - (1996) DANIELS & BRIGHT: /a/ /ɛ/ /i/ /o/ /u/  
  - (1997) BORG:             /a/ /e/ /i/ /o/ /u/  
  - (2009) HUME ET AL.:      /ɐ/ /ɛ/ /i/ /ɔ/ /u/  
  - (2016) PUECH:            /a/ /ɛ/ /ɪ/ /ɔ/ /ʊ/
    * (NB: @MtHume2009 only uses phonetic transcriptions, rather than phonemic, so it is ambiguous whether her choices of vowels reflect phonemes rather than realization.)

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High    "i  iː"       ""         "u  uː"
""        "ɪː"        ""           ""
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/aj/, /ej/, /oj/, /aw/, /ew/, /iw/, /ow/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                   Comment
"a"            "/a/"             "some authors use /ɐ/"
"b"            "/b/"                      ""
"ċ"            "/tʃ/"                     ""
"d"            "/d̪/"                      ""
"e"            "/e/"             "some authors use /ɛ/"
"f"            "/f/"                      ""
"ġ"            "/dʒ/"                     ""
"g"            "/ɡ/"                      ""  
"h"            "/ː/"         "lengthens associated vowels"  
"ħ"            "/ħ/"   "realized by some speakers as [x] or [h]"  
"ie"           "/ɪː/"           "some authors use /ie/"  
"j"            "/j/"                      ""
"k"            "/k/"                      ""
"l"            "/l/"                      ""
"m"            "/m/"                      ""
"n"            "/n/"                      ""
"o"            "/o/"             "some authors use /ɔ/"  
"p"            "/p/"                      ""
"q"            "/ʔ/"                      ""
"r"            "/r/"                      ""
"s"            "/s/"                      ""
"t"            "/t̪/"                      ""
"u"         "/u/, /uː/"          "some authors use /ʊ/"
"v"            "/v/"                      ""
"w"            "/w/"                      ""
"x"            "/ʃ/"         "/ʒ/ appears in certain loanwords"  
"ż"            "/z/"                      ""
"z"            "/ts/"       "/dz/ appears in certain loanwords"
**Digraph**     ""                        ""
"għ"           "/ː/"         "lengthens associated vowels"
"i"         "/i/; /iː/"                   ""
"aj"          "/aj/"                      ""
"aw"          "/aw/"                      ""
"ew"          "/ew/"         "occasionally realized as [ow]"  
"għi"         "/ej/"                      ""
"għu"         "/ow/"                      ""
"iw"          "/iw/"                      ""
"oj"          "/oj/"                      ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Stress

* In words without affixes, Maltese stresses the last VCC or VːC cluster in the word; in words without those clusters, stress falls on the penultimate syllable.
* In loanwords from English, Maltese typically retains the English stress pattern; in loanwords from Italian, the stress pattern is variable, but most commonly based on the Italian vowel length.

# Lenition Processes

* shortening of word-final long vowels
  * stiednu - [st̪ɪːd̪nu] - "they invited" vs.  
    stiednux - [st̪ɪːd̪nuːʃ] - "they did not invite" (Borg, p. 267)
* devoicing of word-final obstruents (Borg, p. 250)
  * kiteb - [kit̪ep] - "he has written"
  * bergħud - [berowt̪] - "flea"
* <għ> represents the Old Arabic phones [ʕ] or [ɣ], which were deleted at some point during the evolution of modern Maltese (Borg, p. 261)
* <h> represents the Old Arabic phone [h], which was deleted at some point during the evolution of modern Maltese

# More Rules

* Clusters of obstruents assimilate to be either fully voiced or fully voiceless
* Short and long vowels are contrastive; likewise short and long consonants
* Word-final long consonants are sometimes shortened
* In the definite article, <il->, the /l/ typically assimilates to subsequent coronal segments (except /d͡ʒ/)
* Metathesis of the definite article occurs before Italian-derived words with initial sibilants
* The <h> in the suffixed pronouns (<-ha>, <-hom>) is replaced by a glide after stressed stem-final vowels
* Word-final vowels lengthen; word boundaries shorten them again, but consonantal suffixes retain the long vowel
* Elision of short vowels occurs in unstressed word-initial syllables
* /i/ is inserted before word-initial CC clusters beginning with sonorants
* CCC clusters are pronounced as CVCC when the middle C is a sonorant
* Glides are inserted between adjacent vowels
* V[+long]C[+glide] → V[-long]C[+glide]C[+glide]  /  _
* Syllabic fusion: /'aːo/ → /'oː/

# References
