---
title: "Greek"
author: "Shiying Yang"
bibliography: el.bib
output: html_document
---

Last-updated: 2019-04-29

# Background

**Language family**: Indo-European / Hellenic/ Greek

* The adopted variety is Standard Modern Greek.
* Speakers reside in a multitude of places, including Albania, Macedonia, Montenegro, Cyprus, Greece, and Ukraine.

# Phonology

## Consonants


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Dental    Alveolar   Velar
Stops                       "p  b"   "t̪  d̪"       ""      "k  ɡ"
Fricatives                  "f  v"   "θ  ð"     "s  z"    "x  ɣ"
Nasals                        "m"      "n̪"        ""        ""
Flaps                         ""       ""         "ɾ"       ""
Approximants                  ""       ""         "l"       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "ε"     ""        "o"
Low    ""      "ɐ"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Uppercase"  "Lowercase"    ""
Α,α,ά  
Β,β  
Γ,γ   
Δ,δ  
Ε,ε,Έ,έ    
Ζ,ζ  
Η,η,Ή,ή  
Θ,θ  
Ι,ι,Ί,ί,Ϊ,ϊ,Ḯ,ΐ  
Κ,κ  
Λ,λ   
Μ,μ  
Ν,ν  
Ξ,ξ  
Ο,ο,Ό,ό  
Π,π  
Σ,ς/σ  
Τ,τ  
Υ,υ,Ύ,ύ,Ϋ,ϋ,Ϋ́,ΰ      
Φ,φ  
Χ,χ  
Ψ,ψ  
Ω,ω,Ώ,ώ
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("Grapheme" = 2, "Phoneme" = 1))
```  

## Correspondence between orthography and phonemes
* Most consonant mappings are shown in table \ref{tab:el_con}, here are mappings for vowels, as well as consonant combinations that do not correspond to single phonemes.
* Stressed vowels (shown by stress markings as seen in ά,έ,ή,ί,ό,ύ,ώ) are represented as the ipa symbol followed by a "1". e.g. /i1/  

|orthography|pronunciation|
|-----------|-------------|
|ξ|/ks/|
|ψ|/ps/|
|α|/ɐ/|
|ά|/ɐ1/|
|ω/ο|/o/|
|ώ/ό|/o1/|
|ι/ϊ/η/υ/ϋ/οι/ει|/i/|
|ί/ΐ/ή/ύ/ΰ/οί/εί|/i1/
|ε/αι|/ε/|
|έ/αί|/ε1/|
|ου|/u/|
|ού|/u1/|
|αυ|/ɐf/|
|αύ|/ɐ1f/|
|ευ|/εv/|
|εύ|/ε1v/|

# Rules

## Issues and representative phonological processes
1. All but two diacritics are kept in the Standard Modern Greek
- stress marking
- umlaut
2. Palatalization of velar consonants
- Velar consonants /k/, /g/, /x/, /ɣ/ surface as palatal allophones [c], [ɟ], [ç], [ʝ] in front of high vowels /i/ and /ε/
- [+velar, +consonantal] -> [+palatal]/__[+high, -consonantal]{\$ or [-back, -consonantal]}
- [+velar, +consonantal][+high, -consonantal] -> [+palatal]/__[+back, -consonantal]

## Lenition Processes

# References {-}
