---
title: "Haitian Creole"
author: "Justin Bai"
bibliography: ht.bib
output: html_document
---
# Background

**Language family**: French Creole

* Speakers of Haitian Creole reside in a multitude of places, including, but not limited to, Haiti, Canada, the United States, France, the Dominican Republic, Cuba, and the Bahamas.

# Phonology

## Consonants

* /ɥ/ is a widely debated phoneme in Haitian Creole, as it only appears in succession with /i/ and can appear in free variation with /w/ [@pindziak, pp. 3-4]. However, I have chosen to include it here in the phonemic inventory.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar   Postalveolar   Palatal   Velar      Glottal
Stops                      "p  b"     "t  d"         ""           ""      "k  ɡ"       ""
Fricatives                 "f  v"     "s  z"       "ʃ  ʒ"         ""       "ɣ"         "h"  
Nasals                       "m"        "n"          ""           ""       "ŋ"         ""
Approximants                 "w"        "l"          ""          "j ɥ"      ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both palatal approximants are voice, but the one on the right is labialized.", general_title = "")
```

## Vowels

* Haitian Creole makes a distinction between oral and nasal vowels. Nasal vowels are orthographically represented with ⟨n⟩ after the vowel [@Faraclas, p. 97].
* Nasal vowels may also be indicated when preceding two ⟨n⟩s [@Faraclas, p. 101].
* /an/, /en/, /on/, and /un/ appear as vowel-consonant sequences when followed by a vowel.
* The grave accent on vowels is also used to indicate non-nasality of vowels when followed by ⟨n⟩ [@Faraclas, p. 101].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""         "u  ũ"
Mid       "e"         ""         "o  õ"
""      "ɛ  ɛ̃"        ""          "ɔ"
Low       ""        "a  ã"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"           "/a/"
"b"           "/b/"
"d"           "/d/"
"e"           "/e/"
"è"           "/ɛ/"
"f"           "/f/"
"g"           "/ɡ/"
"h"           "/h/"
"i"           "/i/"
"j"           "/ʒ/"
"k"           "/k/"
"l"           "/l/"
"m"           "/m/"
"n"           "/n/"
"o"           "/o/"
"ò"           "/ɔ/"
"p"           "/p/"
"r"           "/ɣ/"
"s"           "/s/"
"t"           "/t/"
"v"           "/v/"
"w"           "/w/"
"y"           "/j/"
"z"           "/z/"
**Digraph**    ""
"ch"          "/ʃ/"
"ng"          "/ŋ/"
"ou"          "/u/"
"ui"          "/ɥi/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```
# References
