---
title: "Malayalam"
author: "Becky Mathew"
bibliography: ml.bib
output: html_document
---

# Background

**Language family**: Dravidian, Southern, Tamil-Kannada, Tamil-Kodagu, Tamil-Malayalam, Malayalam

* Malayalam, the official language of Kerala and Lakshadweep, is a Dravidian language from South India. It is also heavily influenced by Sanskrit, and in the modern day, has become influenced by English (introduction of /f/).
* It is spoken in the states of Karnataka, Kerala, Lakshadweep, Puducherry, and Tamil Nadu located in Southern India.

# Phonology

## Consonants

* Stops have a four-way contrast (voiceless unaspirated, voiceless aspirated, voiced unaspirated, voiced aspirated) [@TEMU].
* Consonant length is contrastive [@Jiang2010].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"       Labial        Dental     Alveolar    Retroflex   Postalveolar   Palatal       Velar       Glottal
"Stops (plain)"                  "p  b"        "t̪  d̪"        ""        "ʈ  ɖ"         ""          "ɟ"         "k  ɡ"         ""
"Stops (asp.)"                  "pʰ  bʱ"      "t̪ʰ  d̪ʱ"       ""        "ʈʰ  ɖʱ"       ""          "ɟʱ"        "kʰ  ɡʱ"       ""
Affricates                       ""            ""          ""           ""        "tʃ tʃʰ"       ""            ""          ""
Fricatives                       "ʋ"           ""          "s"          "ʂ"         "ʃ"          ""            ""          "h"
Nasals                           "m"           "n̪"         "n"          "ɳ"         ""           "ɲ"           "ŋ"         ""
Flaps                            ""            ""          "ɾ"          ""          ""           ""            ""          ""
Trills                           ""            ""         "r̥  r"        ""          ""           ""            ""          ""
Approximants                     ""            ""         "l̥  l"      "ɻ  ɭ"        ""           "j"           ""          ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both retroflex approximants are voiced, but the one on the right is lateral.", general_title = "")
```

## Vowels

* Vowel length is contrastive. The diphthongs do not have a length contrast [@Mohanan1986].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /au"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```   

## Alphabet

* Malayalam is written left to right, and does not contain uppercase/lowercase distinctions.
* A consonant letter carries an inherent /a/ or schwa after it, unless indicated by a diacritic.
* Sometimes letters will combine with other letters or diacritics to form a single shape, but they are usually written separately.

Initial vowels and diacritics  
* /a/ = അ  
* /a ː/ = ആ, ാ  
* /i/ = ഇ, ി  
* /i ː/ = ഈ, ീ  
* /e/ = എ,  െ  
* /e ː/ = ഏ, േ  
* /u/ = ഉ, ു  
* /u ː/ = ഊ, ൂ  
* /ai/ = ഐ, ൃ  
* /o/ = ഒ, ൊ  
* /o ː/ = ഓ,ോ  
* /au/ = ഔ, ൌ, ൗ    
* /r̥ i/ = ഋ  
* /r̥ i ː/ = ൠ  
* /l̥ i/ = ഌ, ൢ  
* /l̥ i ː/ = ൡ, ൣ  
* /∅/ = ് (this letter is called a virama, and itremoves the vowel before)  

Consonants with an inherent /a/ following
* /k a/ = ക  
* /kʰ a/ = ഖ  
* /g a/ = ഗ  
* /gʱ a/ = ഘ  
* /ŋ a/ = ങ  
* /tʃ a/ = ച  
* /tʃʰ a/ = ഛ  
* /ɟ a/ = ജ  
* /ɟʱ a/ = ഝ  
* /ɲ a/ = ഞ  
* /ʈ a/ = ട  
* /ʈʰ a/ = ഠ  
* /ɖ a/ = ഡ  
* /ɖʱ a/ = ഢ  
* /ɳ a/ = ണ  
* /t̪ a/ = ത  
* /t̪ʰ a/ = ഥ  
* /d̪ a/ = ദ  
* /d̪ʱ a/ = ധ  
* /n̪ a/ = ന  
* /n a/ = ഩ  
* /p a/ = പ  
* /pʰ a/ = ഫ  
* /b a/ = ബ  
* /bʱ a/ = ഭ  
* /m a/ = മ  
* /j a/ = യ  
* /ɾ a/ = ര  
* /r a/ = റ  
* /l a/ = ല  
* /ɭ a/ = ള  
* /ɻ a/ = ഴ  
* /ʋ a/ = വ  
* /ʃ a/ = ശ  
* /ʂ a/ = ഷ  
* /s a/ = സ  
* /h a/ = ഹ  

Consonants without an inherent vowel following
These kinds of consonants are called chillu or chilluksharam
* /m/ = ം (this letter is called a anusvaram)  
* /h/ = ഃ (this letter is called a visarga)  
* /ɳ/ = ൺ  
* /n̪/ = ൻ  
* /r/ = ർ  
* /l/ = ൽ  
* /ɭ/ = ൾ  
* /k/ = ൿ  
